Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-01-06 supplement.tex
fix up fig label

\section{Gene genealogy reconstruction}

@@

Following \citet{Speidel2019-nj} we also identified “deep branches” within gene
trees. Such a “deep branch” is a branch within a marginal tree (meaning, the
tree at a given locus) that has its upper end (i.e., the node of its
coalescence with another branch) extending to more than 1 million years in age,
and we partitioned deep branches based on their lower node ages (where two or
more branches coalesce into this single branch) into bins between 0 and 1Ma.
Such branches can be categorized by their association with Neanderthals by
comparing mutations that fall upon such a branch to the allele found in a
Neanderthal genome sequence. For this, we used the published high-coverage
Vindija Neanderthal \citep{Prufer2017-kk}. Again following the analysis in
\citet{Speidel2019-nj}, if one or more mutations on a deep branch are shared
with the Neanderthal sequence, the branch is inferred to have passed through
the Neanderthal lineage.

@@

We used \Relate to reconstruct gene genealogies from the simulations under our
four inferred demographic models. From each of these, following the same
approach as for the analysis of the actual data, we computed coalescence rates
within and between sampled populations. From the coalescence rates, we
estimated the inverse instantaneous coalescence rate (IICR, often interpreted
as a trajector of $N_e$) for each population, as well as relative cross
coalescence rates between pairs of populations.

@@

\aprcomment{TODO}

@@

\aprcomment{TODO}

@@

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-f4s-single-origin.pdf}
    \caption{
        \textbf{Predicted \bm{$f_4$} between pairs of modern populations
        and ancient branches from the single-origin model.}
            \aprcomment{The y label should be $f_4^2/f_2f_2$ in these figures!}
    }
    \label{fig:supp-f4s-single-origin}
\end{figure}

>>>>

\section{Gene genealogy reconstruction}
\label{sec:relate}

@@

Following \citet{Speidel2019-nj} we also identified “deep branches” within gene
trees. Such a “deep branch” is a branch within a marginal tree (meaning, the
tree at a given locus) that has its upper end (i.e., the node of its
coalescence with another branch) extending to more than 1 million years in age,
and we partitioned deep branches based on their lower node ages (where two or
more branches coalesce into this single branch) into bins between 0 and 1Ma.
Such branches can be categorized by their association with Neanderthals by
comparing mutations that fall upon such a branch to the allele found in a
Neanderthal genome sequence. For this, we used the published high-coverage
Vindija Neanderthal \citep{Prufer2017-kk}. Again following the analysis in
\citet{Speidel2019-nj}, if one or more mutations on a deep branch (carrying at
least two derived mutations) are shared with the Neanderthal sequence, the
branch is inferred to have passed through the Neanderthal lineage.

@@

We used \Relate \citep{Speidel2019-nj} to reconstruct gene genealogies from the
simulations under our four inferred demographic models. From each of these,
following the same approach as for the analysis of the actual data, we computed
coalescence rates within and between sampled populations. From the coalescence
rates, we estimated the inverse instantaneous coalescence rate (IICR, often
interpreted as a trajector of $N_e$) for each population, as well as relative
cross coalescence rates between pairs of populations.

@@

Following \citet{Speidel2019-nj}, we also characterized Neanderthal-matching
deep branches (see Section~\ref{sec:relate}) in our simulated data under the
four inferred demographic models. Comparing deep-branch distributions from the
data (Fig.~\ref{fig:supp-deep-branches-data}) to the models
(Fig.~\ref{fig:supp-deep-branches-sim}), we find consistent trends between each
of the four models and the data in the ordering of proportions of lineages from
each population that are labeled as Neanderthal-matching deep branches. This
proportion is slightly over-estimated in each of the African populations
compared to the data, but is closest in the merger-with-stem-migration model
(which also fit the LD statistics the best of the four models). Given the
discrepancies between the data IICR and the model IICR, these differenes are
not surprising \aprcomment{what more to say here\dots}

@@

Recent studies \aprcomment{have been quite interested in characterizing the
``earliest divergence time among human populations''}, and such analyses
typically fit simple isolation-with-migration (IM) or ``clean split'' models to
observed allele frequencies or inferred coalescence times among and between
populations \citep{Bergstrom2021-iw} \aprcomment{and others}. Simple IM models
typically fit a constant $N_e$ within each diverged population, their split
time, and a symmetric continuous migration rate (which could be fixed to zero,
equivalent to the ``clean split'' model). We explored the effect of fitting
such simplified models of history to data from more complex models by
simulating the site frequency spectrum (SFS) under each of our four models for
pairs of populations and reinferring an IM and clean split model for each pair.
The simulations assumed 10 diploid individuals were sampled from each
population.

As expected, the clean split models consistently provide divergence times $T$
that are more recent than the IM models. Because $T$ and subsequent migration
can be inversely confounded in SFS inference, fixing migration rates to zero
results in more recent inferred $T$ (Fig.~\ref{fig:supp-IM-misspecification}).
Even though each of our four models inferred divergences of all human
populations $\sim$120kya, ancient structure biases inferences and gives
inflated estimates of $T$. A similar effect occurs when the ancestral
population size increases prior to the population split
\citep{Momigliano2021-th}, which is the case in our inferred single-origin
model. \aprcomment{I want to highlight this more in the main text.} Inferred
$T$ between the Nama and other populations were $>$200kya under the IM model,
and $\sim$50--130kya under the clean split model.

\aprcomment{There's plenty more we could say about this.}

@@

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-f4s-single-origin.pdf}
    \caption{
        \textbf{Predicted \bm{$f_4$} between pairs of modern populations
        and ancient branches from the single-origin model.}
    }
    \label{fig:supp-f4s-single-origin}
\end{figure}
